MS_data_clean

### 2021-11-29 MS_data_modeling/MS_data_explore.Rmd
modeling updates, family level data transofmation on MS data via cat of life

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lifecycle)
library(rvest)
library(dplyr)
library(tidyr)
library(plotly)
library(dplyr)
library(data.table)
library(stringr)
library(readxl)
library(kableExtra)

MS_data <- readxl::read_xlsx('/Users/nickbachelder/Desktop/Fall 2021/Data for Bees/DATA_pollen-diet-breadth_list_09-26-21_ALR_seltmann-colab.xlsx', sheet = 2)

globee <- read.csv('/Users/nickbachelder/Desktop/Bee-Specialization-Modeling/modeling_data/BeeCiteRaw') %>%
  select(-west_specialist, -east_specialist, -cent_specialist, -specialist)

globee_5cut <- read.csv('/Users/nickbachelder/Desktop/Bee-Specialization-Modeling/modeling_data/BeeCiteC5') %>%
  select(-west_specialist, -east_specialist, -cent_specialist, -specialist)

globee_T <- read.csv('/Users/nickbachelder/Desktop/Bee-Specialization-Modeling/modeling_data/BeeCiteW') %>%
  select(-west_specialist, -east_specialist, -cent_specialist, -specialist)

fowler <- read.csv('/Users/nickbachelder/Desktop/Bee-Specialization-Modeling/modeling_data/BeeCiteRaw') %>%
   mutate(interestBee = str_to_title(interestBee)) %>%
  select(interestBee, west_specialist, east_specialist, cent_specialist, specialist)
```

```{r}
MS_data_format <- MS_data %>% mutate(interestBee = str_to_title(paste(Genus, sub(".*? ", "", Species)))) %>% 
  rename(degreeMS = 'Pollen host family #', specializationMS = 'Redefined pollen diet breadth', pollenHosts = 'Pollen hosts') %>%
  select(interestBee, pollenHosts, degreeMS, specializationMS, References) %>%
  filter(!is.na(degreeMS)) %>% arrange(interestBee) %>% mutate(pollenHosts = str_to_title(gsub("[A-Za-z]*[^A-Za-z/, ]\\S*\\s*", "", pollenHosts)))

MS_data_format
```

```{r}
globee_format <- globee %>% mutate(interestBee = str_to_title(interestBee)) %>% rename(degreeGLOBEE = degree) %>% 
  select(interestBee, degreeGLOBEE) %>% unique() %>%
  arrange(interestBee)

globee_5cut_format <- globee_5cut %>% mutate(interestBee = str_to_title(interestBee)) %>% rename(degreeGLOBEE_C5 = degree) %>% 
  select(interestBee, degreeGLOBEE_C5) %>% unique() %>%
  arrange(interestBee)

globee_T_format <- globee_T %>% mutate(interestBee = str_to_title(interestBee)) %>% rename(degreeGLOBEE_T = degree) %>% 
  select(interestBee, degreeGLOBEE_T) %>% unique() %>%
  arrange(interestBee)
```

@@

compare_df <- compare_df %>% merge(fowler %>% rename(specialistFOWLER = specialist) %>% 
                                     select(interestBee, specialistFOWLER), by = c('interestBee'), all = T) %>% unique() %>%
  mutate(existsGLOBEE = ifelse(is.na(degreeGLOBEE), 0, 1), existsMS = ifelse(is.na(degreeMS), 0, 1), 
         existsFOWLER = ifelse(is.na(specialistFOWLER), 0, 1),
         existsALL = existsGLOBEE*existsMS*existsFOWLER)
```

@@

data.frame(data_sources, counts) %>% mutate('Percent of Data' = counts / nrow(compare_df)) %>% 
  rename('Data Source' = data_sources, 'Unique Genus Count' = counts) %>% kbl(caption = "Bee Data Sources Summary") %>%
  kable_styling()
```

```{r}
compare_df %>% filter(existsALL == 1) %>% group_by(specializationMS) %>% summarize('Mean GLOBI Degree'  = mean(degreeGLOBEE),
                                                                                   'GLOBI Degree IQR' = paste('[', round(quantile(degreeGLOBEE, 0.25),
                                                                                                                         2), 
                                                                                                                  ',', round(quantile(degreeGLOBEE, 0.75), 2),
                                                                                                            ']'),
                                                                                   'Mean GLOBI C5 Degree' = mean(degreeGLOBEE_C5),
                                                                                   'GLOBI C5 Degree IQR' = paste('[', round(quantile(degreeGLOBEE_C5, 0.25), 2),
                                                                                                                  ',', round(quantile(degreeGLOBEE_C5, 0.75), 2),
                                                                                                            ']'),
                                                                                   'Mean GLOBI T Degree' = mean(degreeGLOBEE_T),
                                                                                   'GLOBI T Degree IQR' = paste('[', round(quantile(degreeGLOBEE_T, 0.25), 2),
                                                                                                                  ',', round(quantile(degreeGLOBEE_T, 0.75), 2),
                                                                                                            ']'),
                                                                                   'Mean MS Degree' = mean(degreeMS),
                                                                                   'MS Degree IQR' = paste('[', round(quantile(degreeMS, 0.25), 2), 
                                                                                                                  ',', round(quantile(degreeMS, 0.75), 2),
                                                                                                            ']'),
                                                                                   'Percent Fowler Specialist' = mean(specialistFOWLER)
) %>% kbl(caption = 'Plant Family Degrees by MS Specialization') %>%
  kable_styling()
```

@@

compare_df %>% filter(specialistFOWLER == 1, specializationMS  == 'Polylectic') %>% 
  select(interestBee, specializationMS, specialistFOWLER, degreeGLOBEE_C5, degreeGLOBEE)
```

@@

compare_df %>% filter(specialistFOWLER == 0, specializationMS  != 'Polylectic') %>% 
  select(interestBee, specializationMS, specialistFOWLER, degreeGLOBEE_C5, degreeGLOBEE) %>% filter(degreeGLOBEE_C5 > 5)
```

## Plan: want to add pollen hosts so I can filter on pollen family. Discuss how to parse the authors out of appropriate MS column. ## DONE

@@

## Create slides / tables explaining odd bees between datasets (ie. fowler specialist vs MS non-specialist and vis versa)

## Check how many GLOBI bees were cut when filtering for fowler, add them back (update previously done tables)

```{r}

```

```{r}

```

```{r}

```

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lifecycle)
library(rvest)
library(dplyr)
library(tidyr)
library(plotly)
library(dplyr)
library(data.table)
library(stringr)
library(readxl)
library(kableExtra)
library(rcol)
library(plyr)

MS_data <- readxl::read_xlsx('/Users/nickbachelder/Desktop/Fall 2021/Data for Bees/DATA_pollen-diet-breadth_list_10-26-21_ALR seltmann-colab.xlsx', sheet = 2)

globee <- read.csv('/Users/nickbachelder/Desktop/Bee-Specialization-Modeling/modeling_data/BeeCiteRaw') %>%
  dplyr::select(-west_specialist, -east_specialist, -cent_specialist, -specialist)

globee_5cut <- read.csv('/Users/nickbachelder/Desktop/Bee-Specialization-Modeling/modeling_data/BeeCiteC5') %>%
  dplyr::select(-west_specialist, -east_specialist, -cent_specialist, -specialist)

globee_T <- read.csv('/Users/nickbachelder/Desktop/Bee-Specialization-Modeling/modeling_data/BeeCiteW') %>%
  dplyr::select(-west_specialist, -east_specialist, -cent_specialist, -specialist)

fowler <- read.csv('/Users/nickbachelder/Desktop/Bee-Specialization-Modeling/Fowler-GLOBI Modeling/fowler') %>%
   mutate(interestBee = str_to_title(interestBee)) %>%
  dplyr::select(interestBee, west_specialist, east_specialist, cent_specialist, specialist) %>% 
  filter(grepl('^\\w+\\s\\w+$', interestBee))
```

```{r}
MS_data_format <- MS_data %>% mutate(interestBee = str_to_title(paste(Genus, sub(".*? ", "", Species)))) %>% 
  dplyr::rename(degreeMS = 'Pollen host family #', specializationMS = 'Redefined pollen diet breadth', pollenHosts = 'Pollen hosts') %>%
  dplyr::select(interestBee, pollenHosts, degreeMS, specializationMS, References) %>%
  filter(!is.na(specializationMS)) %>% arrange(interestBee) %>% 
  filter(grepl('^\\w+\\s\\w+$', interestBee))

MS_data_format %>% filter(!(is.na(specializationMS)), is.na(pollenHosts)) ### there
```

### create family level column by pulling

```{r}
cp_nu_search(q='Sesamum', status = "accepted")$result$classification
```

```{r}
MS_data_format$pollenHosts <-  gsub("\\?.*", "", MS_data_format$pollenHosts)
MS_data_format$pollenHosts <-  gsub("\\..*", "", MS_data_format$pollenHosts)

ms_pollen_hosts = na.omit(unique(

  sub("\\/.*", "", word(unlist(str_split(unique(MS_data_format$pollenHosts), pattern = ', ')), 1))

  ))

## pulls all unique maximum level taxonomic names from MS pollen hosts
ms_pollen_hosts <- ms_pollen_hosts[!(ms_pollen_hosts == '')]

ms_pollen_hosts
```

```{r}
library(rcol)
library(pbapply)

find_family <- function(name) {
  try <- try(cp_nu_search(q=name, status = "accepted"), silent=TRUE)
  if ("try-error" %in% class(try)) {
    label <- 'Unknown'
  }
  try <- try(str_to_title(tolower(
    (rbind( data.frame(try$result$classification[1]),
        data.frame(try$result$classification[2]),
       data.frame(try$result$classification[3])) %>% filter(rank == 'family'))[1,]$label
    )), silent=TRUE)
  if ("try-error" %in% class(try)) {
    label <- 'Unknown'
  }
  else {
    label <- try
  }
  return(label)
}

familys <- pblapply(ms_pollen_hosts, find_family)
```

```{r}
familys <- unlist(familys)
family_key <- data.frame(cbind(ms_pollen_hosts, familys)) %>% `colnames<-`(c('max_taxons', 'pollen_host_familys'))

family_key
```

## Make the below two entries have 0 rows with Avery's help
```{r}
family_key %>% filter(is.na(pollen_host_familys))
```

```{r}
family_key %>% filter(pollen_host_familys == 'Unknown', grepl( ',', max_taxons, fixed = TRUE))
```

## make sure the below has no family name synonyms. Take out the unknownand NA values until we fix them with avery.

```{r}
pollen_family_MS <- MS_data_format %>% select(interestBee, pollenHosts) %>% separate_rows(pollenHosts, sep=', ') %>% mutate(max_taxons = word(pollenHosts, 1)  ) %>% left_join(family_key, by = 'max_taxons') %>% select(interestBee, pollen_host_familys) %>% 
  group_by(interestBee) %>% 
  dplyr::summarise(pollen_host_familys = paste(unique(as.vector(pollen_host_familys)), collapse = ', '))

MS_data_format <- MS_data_format %>% left_join(pollen_family_MS, by = 'interestBee') 
```

```{r}
avery_send <- MS_data_format %>% select(interestBee, pollenHosts, pollen_host_familys) %>% mutate(ifelse(pollen_host_familys == 'NA', NA, pollen_host_familys))

write.csv(avery_send, "/Users/nickbachelder/Desktop/Bee-Specialization-Modeling/Family_Level_Transform.csv")

avery_send
```

### Clean GLOBEE data

```{r}
pollenHostGLOBEE <- aggregate(interact~interestBee, globee_5cut, FUN=toString) %>% mutate(interestBee = str_to_title(interestBee)) %>% 
  dplyr::rename('pollenHostGLOBEE' = interact) %>% 
  filter(grepl('^\\w+\\s\\w+$', interestBee))
## this is plant interactions for each bee genus according to globi

globee_format <- globee %>% mutate(interestBee = str_to_title(interestBee)) %>% dplyr::rename(degreeGLOBEE = degree) %>% 
  dplyr::select(interestBee, degreeGLOBEE, n_Bee) %>% unique() %>%
  arrange(interestBee) %>% merge(pollenHostGLOBEE, by = 'interestBee') %>% 
  filter(grepl('^\\w+\\s\\w+$', interestBee))

globee_5cut_format <- globee_5cut %>% mutate(interestBee = str_to_title(interestBee)) %>% dplyr::rename(degreeGLOBEE_C5 = degree) %>% 
  dplyr::select(interestBee, degreeGLOBEE_C5) %>% unique() %>%
  arrange(interestBee)%>% 
  filter(grepl('^\\w+\\s\\w+$', interestBee))

globee_T_format <- globee_T %>% mutate(interestBee = str_to_title(interestBee)) %>% dplyr::rename(degreeGLOBEE_T = degree) %>% 
  dplyr::select(interestBee, degreeGLOBEE_T) %>% unique() %>%
  arrange(interestBee)%>% 
  filter(grepl('^\\w+\\s\\w+$', interestBee))
```

@@

compare_df <- compare_df %>% merge(fowler %>% dplyr::rename(specialistFOWLER = specialist) %>% 
                                     dplyr::select(interestBee, specialistFOWLER), by = c('interestBee'), all = T) %>% unique() %>%
  mutate(existsGLOBEE = ifelse(is.na(degreeGLOBEE), 0, 1), existsMS = ifelse(is.na(specializationMS), 0, 1), 
         existsFOWLER = ifelse(is.na(specialistFOWLER), 0, 1),
         existsALL = existsGLOBEE*existsMS,
         pollenHostGLOBEE = str_to_title(pollenHostGLOBEE)) %>%
  mutate(specialistFOWLER = ifelse(is.na(specialistFOWLER), 0 , specialistFOWLER))%>% 
  dplyr::rename(pollenHostsMS = pollenHosts ,pollenHostsFamilysMS = pollen_host_familys,ReferencesMS = References, nBeeGLOBEE = n_Bee) %>%
  dplyr::select(interestBee, specializationMS, nBeeGLOBEE, degreeGLOBEE, degreeGLOBEE_C5, degreeGLOBEE_T, specialistFOWLER, 
         existsGLOBEE, existsMS, existsFOWLER, existsALL, pollenHostsMS, pollenHostsFamilysMS, pollenHostGLOBEE, ReferencesMS, degreeMS) %>%
  rowwise() %>%
  dplyr::mutate(pollenHostsFamilysMS = ifelse(pollenHostsFamilysMS == 'NA', NA, pollenHostsFamilysMS) ) %>%
  dplyr::mutate(FamilyDegreeMS = ifelse(!is.na(pollenHostsFamilysMS), length(strsplit(pollenHostsFamilysMS, split = ', ')[[1]]), 0  ) )

compare_df
```

@@

data.frame(data_sources, counts) %>% mutate('Percent of Data' = counts / nrow(compare_df)) %>% 
  dplyr::rename('Data Source' = data_sources, 'Unique Species Count' = counts) %>% kbl(caption = "Bee Data Sources Summary") %>%
  kable_styling()
```

```{r}
compare_df %>% filter(existsALL == 1) %>% group_by(specializationMS) %>% dplyr::summarize('Mean GLOBI Degree'  = mean(degreeGLOBEE),
                                                                                   'GLOBI Degree IQR' = paste('[', round(quantile(degreeGLOBEE, 0.25),
                                                                                                                         2), 
                                                                                                                  ',', round(quantile(degreeGLOBEE, 0.75), 2),
                                                                                                            ']'),
                                                                                   'Mean GLOBI C5 Degree' = mean(degreeGLOBEE_C5),
                                                                                   'GLOBI C5 Degree IQR' = paste('[', round(quantile(degreeGLOBEE_C5, 0.25), 2),
                                                                                                                  ',', round(quantile(degreeGLOBEE_C5, 0.75), 2),
                                                                                                            ']'),
                                                                                   'Mean GLOBI T Degree' = mean(degreeGLOBEE_T),
                                                                                   'GLOBI T Degree IQR' = paste('[', round(quantile(degreeGLOBEE_T, 0.25), 2),
                                                                                                                  ',', round(quantile(degreeGLOBEE_T, 0.75), 2),
                                                                                                            ']'),
                                                                                   'Percent Fowler Specialist' = mean(specialistFOWLER)
) %>% kbl(caption = 'Plant Family Degrees by MS Specialization') %>%
  kable_styling()
```

@@

odd_generalists <- compare_df %>% filter(specialistFOWLER == 1, specializationMS  == 'Polylectic') %>% 
  dplyr::select(interestBee, specializationMS, specialistFOWLER, degreeGLOBEE_C5, degreeGLOBEE, pollenHostsFamilysMS, pollenHostsMS, pollenHostGLOBEE, ReferencesMS) %>% filter(degreeGLOBEE_C5 == 1)

odd_generalists

write.csv(odd_generalists, '/Users/nickbachelder/Desktop/Bee-Specialization-Modeling/modeling_data/odd_polylectic.csv')
```

@@

odd_specialists <- compare_df %>% filter(specialistFOWLER == 0, specializationMS  != 'Polylectic') %>% 
  dplyr::select(interestBee, specializationMS, specialistFOWLER, degreeGLOBEE_C5, degreeGLOBEE, pollenHostsMS, pollenHostsFamilysMS, pollenHostGLOBEE, ReferencesMS) %>% filter(degreeGLOBEE_C5 > 3)

odd_specialists

write.csv(odd_specialists, '/Users/nickbachelder/Desktop/Bee-Specialization-Modeling/modeling_data/odd_specialists.csv')
```

```{r}
# Bees NOT in MS

missing_MS <- compare_df %>% filter(existsMS == 0)

missing_MS

write.csv(missing_MS, '/Users/nickbachelder/Desktop/Bee-Specialization-Modeling/modeling_data/missing_MS.csv')
```

## Plan: want to add pollen hosts from MS data so I can filter on pollen family. Discuss how to parse the authors out of appropriate MS column. ## DONE
NEED TO DISCUSS: The MS pollen hosts are a mix of genus and families, randomly. Maybe we can ask MS to seperate these into two columns, or just simply make a family level column only.

@@

## Create slides / tables explaining odd bees between datasets (ie. fowler specialist vs MS non-specialist and vis versa) ## WIP

## Check how many GLOBI bees were cut when filtering for fowler, add them back (update previously done tables). ## DONE
DISCUSS: Turns out that we didnt cut globi bees, but we instead cut FOWLER bees. I unfiltered the fowler bees and it fixed the ratios. Now we can actually discuss data proportions

### Also: Added the target pollen hosts from globi so we can compare to pollen hosts from MS
This way we can directly compare the MS pollen hosts to the GLOBI pollen hosts. THe only issue is the above problem with a mix of taxonomic levels with the MS data

---------------

For next week:

Check out the web scraper to fix MS names to family level

Pull out data that is only in GLOBI and/or Fowler, use as group to check for old names as reason.

--------------

## Modeling below

# (1) Model same as old, except use conclusions from MS not fowler

```{r}
modeling_df <- compare_df %>% filter(existsMS == 1, existsGLOBEE == 1) %>% mutate(specialist = ifelse(specializationMS == 'Polylectic', 1, 0 )) %>% dplyr::select(interestBee, nBeeGLOBEE, degreeGLOBEE, degreeGLOBEE_C5, degreeGLOBEE_T, FamilyDegreeMS, specialist)

modeling_df
```

```{r}
calc_error_rate = function(predicted.value, true.value){
  return(mean(true.value!=predicted.value))
}
records = matrix(NA, nrow=3, ncol=2)
colnames(records) = c("train.error","test.error")
rownames(records) = c("tree","logistic","lasso")

nfold <- 10

n <- nrow(modeling_df)

in.trn <- sample.int(n, 0.8*n) 
trn.cl <- modeling_df[ in.trn,]
tst.cl <- modeling_df[-in.trn,]

folds <- sample(cut(1:nrow(trn.cl), breaks=nfold, labels=FALSE))
```

```{r}
library(tree)
library(maptree)

tree <- tree(as.factor(specialist) ~ degreeGLOBEE_T + nBeeGLOBEE, data = trn.cl)
cv <- cv.tree(tree, rand = folds, FUN = prune.misclass, K = 10)

best.size.cv = min(cv$size[cv$dev == min(cv$dev)])

tree.pruned <- prune.misclass(tree, best= best.size.cv)

train_pred_tree <- predict(tree.pruned, trn.cl %>% select(degreeGLOBEE_T, nBeeGLOBEE), type = 'class')
test_pred_tree <- predict(tree.pruned, tst.cl %>% select(degreeGLOBEE_T, nBeeGLOBEE), type = 'class')

train_err <- calc_error_rate(train_pred_tree, trn.cl$specialist)
test_err <- calc_error_rate(test_pred_tree, tst.cl$specialist)

1 - train_err
1 - test_err
```

```{r}
glmfit <- glm(as.factor(specialist) ~ degreeGLOBEE_T + nBeeGLOBEE, data = trn.cl, family = 'binomial')
predicttrain_log = predict(glmfit, trn.cl %>% select(degreeGLOBEE_T, nBeeGLOBEE), type="response")
predicttest_log = predict(glmfit, tst.cl %>% select(degreeGLOBEE_T, nBeeGLOBEE), type="response")

predicttrain_log = ifelse(predicttrain_log > 0.5, 1, 0)
predicttest_log = ifelse(predicttest_log > 0.5, 1, 0)

trainerror <- calc_error_rate(predicttrain_log, trn.cl$specialist)
valerror <- calc_error_rate(predicttest_log, tst.cl$specialist)

1 - trainerror
1 - valerror
```

```{r}
library(randomForest)

rf_fit <-  randomForest(as.factor(specialist) ~ degreeGLOBEE_T + nBeeGLOBEE, data = trn.cl, importance=TRUE)

predict_train <- predict(rf_fit, newdata = trn.cl %>% select(degreeGLOBEE_T, nBeeGLOBEE), type="class")
predict_test <- predict(rf_fit, newdata = tst.cl %>% select(degreeGLOBEE_T, nBeeGLOBEE), type="class")

train_err <- calc_error_rate(as.character(predict_train), trn.cl$specialist)

test_err <- calc_error_rate(as.character(predict_test), tst.cl$specialist)

1 - train_err
1 - test_err
```
# Try multi feature with best model:

```{r}
rf_fit <-  randomForest(as.factor(specialist) ~ ., data = trn.cl, importance=TRUE)

predict_train <- predict(rf_fit, newdata = trn.cl %>% select(-specialist), type="class")
predict_test <- predict(rf_fit, newdata = tst.cl %>% select(-specialist), type="class")

train_err <- calc_error_rate(as.character(predict_train), trn.cl$specialist)

test_err <- calc_error_rate(as.character(predict_test), tst.cl$specialist)

1 - train_err
1 - test_err
```

# Begin Baysian Approaches

```{r}
modeling_df
```

```{r}
compare_df %>% filter(interestBee == 'Agapostemon Angelicus') 
```
